The Impact of Alternation

### 2017-05-10 draft.tex
ok

Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
$\Bool$ and $\Data$, respectively, we denote by
$\Form^+(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
$\fv{\Bool}{\phi} \subseteq \vec{b}$, $\fv{\Data}{\phi} \subseteq
\vec{x}$ and each variable from $\vec{b}$ occurs under an even number
of negations.

@@

@@

As pointed out by a number of authors (e.g.\ \cite{mcmillan06}) the
bottleneck of predicate abstraction is the high cost of existential
quantifier elimination required to compute the $\Post{\A}(\phi,a)$
function --- we recall that $\Post{\A}(\phi,a)$ is defined by the
formula $\overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}],a)$. The main idea of the
\impact~ procedure is that interpolation can be used instead
quantifier elimination, because an interpolant over-approximates the
post-images along the counterexample path in a way that is sufficient
to prove its spuriousness.

With this in mind, Algorithm \ref{alg:impact} builds an ART for the
given automaton and uses interpolants, instead of abstract
post-images, to label its nodes. To speedup the computation we use two
additional labeling functions for bookkeeping. Formally, the ART
structure used by Algorithm \ref{alg:impact} is defined as $\Art =
\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
defined as for Algorithm \ref{alg:predabs} and,
moreover: \begin{compactitem}
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
  with time-stamped formulae, in which $Q_i$ may occur under negation,
  and
%
\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
  depth\footnote{We consider that the root node has zero depth.}
  $k\geq0$ in the tree $(N,E)$, then we have:
  \[\begin{array}{rcl}
  T(e) & = & \Theta(n,k) \\
  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
  \end{array}\]
\end{compactitem}
For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
$(n,a,m)$. The idea behind this additional bookkeeping is that, given
a sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path
from the root to a node $n$ in $\Art$, the conjunction: 
\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q
\rightarrow \bot)\] is equivalent to the interpolation problem
$\pi(u)$. We assume furthermore a total alphabetical order $\prec$ on
$\Sigma$ and lift it to the total lexicographical order $\prec^*$ on
$\Sigma^*$. We write $n \prec^* m$ if the sequence labeling the path
from $\rootNode$ to $n$ is strictly smaller, in the lexicographical
order, than the sequence labeling the path from $\rootNode$ to $m$. A
node $n \in N$ is \emph{covered by $m$} if $(n,m) \in \lhd$ and simply
covered when $m$ is not important. A node $n$ is \emph{closed} if for
all nodes $m \in N$ such that $m \prec^* n$, either $\Lambda(n)
\not\models \Lambda(m)$ or $(n,m) \in \lhd$.

@@

  \For{$m \in N$ such that $m \prec^* n$}
  \Comment{close $n$}

@@

@@

  \If{$n$ is not covered}

@@

  \State let $\rootNode,n_1,\ldots,n_k$ be the sequence of nodes from $\rootNode$ to $n$

@@

  \State let $s$ be a fresh node and $e \leftarrow (n,a,s)$

@@

>>>>

Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
$\Bool$ and $\Data$, respectively, we denote by
$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
under an even number of negations.

@@

\paragraph{\em Remark}
While computing the concrete post-image $\Post{\A}$ relies on an
expensive quantifier elimination, it is common belief that this should
hold also for $\AbsPost{\A}$. However, the choice of conjoined
predicates that form $\AbsPost{\A}$ requires only a quantifier-free
satisfiability query: $\Post{\A}(\phi,a) \models \pi \iff \exists
\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
\models \pi \iff \Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
\neg\pi$ unsatisfiable. \qed

@@

As pointed out by a number of authors, the bottleneck of predicate
abstraction is the high cost of reconstructing parts of the ART,
subsequent to the refinement of the set of predicates. The main idea
of the \impact~ procedure \cite{mcmillan06} is that this can be
avoided and the refinement (strenghtening of the node labels of the
ART) can be performed in-place. This refinement step requires an
update of the covering relation, because a node that used to cover
another node might not cover it after the strenghtening of its label.

The ART built by Algorithm \ref{alg:impact} corresponds to an
unfolding of the input automaton by enumerating the input event
sequences in $\Sigma^*$. Each time a new node is created, its label is
initially $\top$. This label is subject to subsequent strenghtening,
when a spurious counterexample is encountered. To speedup the
computation of interpolants for spurious counterexamples, we use two
additional labeling functions for bookkeeping. Formally, the ART
structure used by Algorithm \ref{alg:impact} is defined as $\Art =
\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
defined as for Algorithm \ref{alg:predabs} and,
moreover: \begin{compactitem}
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
  with time-stamped formulae, and
%
\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
  depth\footnote{We consider that the root node has zero depth.}  $k$
  in the tree $(N,E)$, then we have:
  \[\begin{array}{rcl}
  T(e) & = & \Theta(n,k) \\
  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
  \end{array}\]
\end{compactitem}
For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
$(n,a,m)$. This additional bookkeeping is used as follows. Given a
sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
the root to a node $n$ in $\Art$, the conjunction:
\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q
\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.

We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
to the total lexicographical order $\prec^*$ on $\Sigma^*$. We write
$n \prec^* m$ if the sequence labeling the path from $\rootNode$ to
$n$ is strictly smaller, in the lexicographical order, than the
sequence labeling the path from $\rootNode$ to $m$. A node $n \in N$
is \emph{covered by $m$} if $(n,m) \in \lhd$. We say that $n$ is just
\emph{covered} when $m$ is not important. A node $n$ is \emph{closed}
if for all nodes $m \in N$ such that $n \prec^* m$, either $\Lambda(n)
\not\models \Lambda(m)$ or $(n,m) \in \lhd$ holds. To efficiently
check that a node is closed, one may keep all the nodes from the ART
in a list, in preorder.

@@

  \For{$m \in N$ such that $n \prec^* m$ and $n$ is not an ancestor of $m$}
  \Comment{close $n$}

@@

  \State {\bf goto} \ref{ln:not-covered}

@@

  \If{$n$ is not covered}
  \label{ln:not-covered}

@@

  \State let $\rootNode,n_1,\ldots,n_k,n$ be the sequence of nodes from $\rootNode$ to $n$

@@

  \State let $s$ be a fresh node and $e = (n,a,s)$ be a new edge

@@

The execution of Algorithm \ref{alg:impact} consists of three
phases\footnote{These phases correspond to the \textsc{Close},
  \textsc{Refine} and \textsc{Expand} procedures from the original
  description of \impact~\cite{mcmillan06}.}: \emph{close},
\emph{refine} and \emph{expand}.
